---
title: Feasibility of offsetting methane emissions from cows by planting trees.
authors:
  - name: Peter Levy
    affiliation: CEH
    roles: writing
    corresponding: true
bibliography: cows.bib
editor:
  markdown:
    wrap: sentence
format: 
  html:
    self-contained: true
---

```{r render, eval=FALSE, echo = FALSE}
library(quarto)
quarto::quarto_render("qmd/2025-11-11-cows.qmd", output_format = "html")
quarto::quarto_render("qmd/2025-11-11-cows.qmd", output_format = "md")
```

```{r startup, echo = FALSE, warning = FALSE, message = FALSE}
here::i_am("qmd/2025-11-11-cows.qmd")
library(units)
library(readxl)
library(stringr)
library(data.table)
library(ggplot2)
library(ggforce)
```
```{r, echo = FALSE}
GWP_CH4_CO2_100yr <- 25  # from IPCC AR4
# dairy = 100, beef = 50-75
# an additional 20% comes from manure
e_cow <- set_units(75 * 1.2, kg / year)  # ch4 emitted by one cow per year
```

## Question:
*What area of land would need to be planted with trees to make up for the methane 
emissions from grazing cattle?*

This may seem a simple question, but equating the two sides of the equation is
complicated because it is hard to compare like with like. 
Emissions from grazing (adult) cattle are constant over time, and depend only on 
the number of cattle, not the area they occupy.
Uptake by planted woodland is proportional to the area and varies strongly over 
time: initially zero or negative after planting, then positive until a peak 
growth rate after ~30-50 years, then declining to zero as the woodland matures. 
At some point a mature woodland reaches equilibrium and becomes carbon neutral 
(where photosynthesis is balanced by total ecosystem respiration).

So the question posed cannot be answered without specifying a time period over
which to consider the balance between emissions and uptake.
We can rephrase the question in a number of ways, depending on how we choose to
compare the two.

1. What area of land would need to be planted with trees to offset the methane 
emissions from a single cow over the next 25 years (i.e. so as to meet the Net 
Zero 2050 target for that single cow)?
2. What area of land would need to be planted with trees to offset the methane 
emissions from a single cow over the next 100 years?
3. On a single farm with a fixed area, how long could grazing continue at a 
reasonable stocking density before the area of the land needed for woodland 
precludes grazing?

The timing of the woodland planting is also important, so we also need to 
distinguish whether this happens at the start of a project or is continuously 
on-going over time. There is a big benefit to the former.

We can consider each of these questions, given some parameters for emissions 
from cows and uptake by trees.


### Methane from cows
One dairy cow emits about 100 kg of methane per year from enteric fermentation; 
for non-dairy cattle, the value is ~50-75 kg/yr - e.g. see 
[here](https://teagasc.ie/environment/climate-change-air-quality/methane/) 
or [here](https://naei.energysecurity.gov.uk/emission-factors)). 
Methane emissions from manure adds around 20% to this.
Here we use a value of 75 x 1.2 = `r e_cow` kg of CH4 as an average, 
intermediate between beef and dairy.
```{r, echo = FALSE}
e_cow <- e_cow * GWP_CH4_CO2_100yr  # co2 equivalent emitted by one cow per year
n_cows <- 9.493333e6 # 9.5 million cattle population in 2023
```
Methane has a global warming potential about `r GWP_CH4_CO2_100yr` times that of 
CO₂ over a 100-year period, so that is equivalent to `r e_cow` kg of CO₂ 
per year.
The UK has a cattle population of around `r round(n_cows/1e6, 1)` million, so an
annual emission of `r round(set_units(e_cow * n_cows, Tg/yr), 2)` Tg CO$_2$-e/yr.


### Carbon uptake by trees
For a crude comparator, `r e_cow` kg of CO₂ is equivalent to a mature Sitka 
spruce tree (dbh of 40 cm, age ~60 years [@Levy2004]). So we need to add one 
mature tree, requiring 60 years to grow and occupying about 16 m$^2$ 
(4 m spacing), for each cow, every year.

To be properly quantitative so as to answer questions 1-3, we need to consider 
the time course of tree growth in woodlands. Below, we show data for 
sycamore/ash/birch from the [Woodland Carbon Code carbon calculation spreadsheet](https://www.woodlandcarboncode.org.uk/template-documents-and-tools-version-22).
(version 2.4.1, April 2024). We use this as typical of UK planting for 
conservation projects in answering the questions below, but other species could 
be chosen. Results for Sitka spruce, the main commercial species, are rather 
similar, although the peak growth occurs later.

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Figure 1: Carbon stock in sycamore/ash/birch woodland over time, with 2.5 m initial spacing, yield class 8, no thinning."} 
fname <- here::here("data-raw", "WCC_CarbonCalculationSpreadsheet_Version2.4.1_April2024.xlsx")
dt <- readxl::read_excel(fname, sheet = "Biomass Carbon Lookup Table")
setDT(dt)
names(dt) <- make.names(names(dt))
setnames(dt, c("Spacing..m.", "Cumulative.Total.Sequestrn..tCO2e.ha.", "Period..year."), c("Spacing", "S_c", "time"))

# choose species 
dt <- dt[Species == "SAB" & Spacing == 2.5 & Yield.Class == 8 & Management == "NO_thin", .(time, S_c)]
# dt <- dt[Species == "SS" & Spacing == 2.0 & Yield.Class == 12 & Management == "NO_thin", .(time, S_c)]

dt[, time := as.numeric(stringr::str_split_i(dt$time, "-", 2))]
# insert a row at time = 0, S_c = 0
dt <- rbindlist(list(data.table(time = 0, S_c = 0), dt))
dt[, S_c := set_units(S_c, tonnes / ha)]
dt[, S_c := set_units(S_c, kg / m^2)]

dt_time <- data.table(time = seq(0, max(dt$time), by = 1))
dt <- merge(dt_time, dt, by = "time", all.x = TRUE)
dt[, S_c := zoo::na.spline(S_c, x = time, na.rm = FALSE)]
dt[, S_c := set_units(S_c, kg / m^2)] # units get lost - need to reset
dt[, time := set_units(time, yr)]
dt <- dt[-1] # get rid of year 0

p <- ggplot(dt, aes(x = time, y = S_c)) +
  geom_line() +
  xlab("Time (years)") +
  ylab("Carbon stock as CO2") +
  theme_minimal()
p
```

The annual change in carbon stock gives the annual flux of CO₂ into the woodland
shown below. The solid line shows the instantaneous annual flux in that year,
while the dashed line shows the mean flux over the period since planting. 
Given the time period of interest, we can use the latter to compare with the 
constant annual emissions from cows.

```{r, echo = FALSE, fig.cap = "Figure 2: Annual CO$_2$ flux over time following planting sycamore/ash/birch woodland over time, with 2.5 m initial spacing, yield class 8, no thinning. Solid line shows the instantaneous uptake rate for each year; the dashed line shows the mean uptake rate since planting; the dotted line shows the mean uptake rate over time assuming planting is continuous, with an equal area planted each year."}
# get annual flux by difference
dt[, f_c := c(0, diff(S_c))]
dt$f_c[1] <- dt$f_c[2] / 2 # make first value non-zero
dt[, f_c := set_units(f_c, kg / m^2)] # units get lost - need to reset
dt[, f_c_mean := cumsum(f_c) / time]

# find the weighted average flux from the first to the current time
# this is the convolved rate, assuming a continuous affn rate
dt[, w := time]
# dt[, w := 1] # or unweighted
# these give the same result with w = 1
dt[, f_c_mean_w := cumsum(f_c_mean * w) / cumsum(w)]
# should give same answer for a given time period
# dt[time <= set_units(10, yr), weighted.mean(f_c_mean, w)]

p <- ggplot(drop_units(dt), aes(x = time, y = f_c)) +
  geom_line() +
  geom_line(aes(y = f_c_mean), linetype = "dashed") +
  geom_line(aes(y = f_c_mean_w), linetype = "dotted") +
  # geom_point() +
  xlab("Time") +
  ylab("Carbon uptake rate (kg CO2/m^2/yr)") +
  theme_minimal()
p
```

### Q1a. What area of land would need to be planted with trees *in 2025* to offset the methane emissions from a single cow over the next 25 years (i.e. so as to meet the Net Zero 2050 target for that single cow)?
```{r, echo = FALSE, warning = FALSE, message = FALSE}
rotn_length <- 25
# f_c_mean <- mean(dt$S_c[1:rotn_length]) # gives same answer
f_c_mean <- dt[time == rotn_length, f_c_mean]
S_c_25y <- dt[time == rotn_length, S_c]
f_c_mean <- dt[time == rotn_length, S_c / time]
f_c_mean <- set_units(f_c_mean, kg / m^2 / year)
required_affn <- e_cow / f_c_mean  # m2 of woodland needed to offset one cow's emissions per year
required_affn_1a <- required_affn
```
In this scenario, we consider the area of woodland planted in 2025 needed to 
offset the methane emissions from a single cow over the following `r rotn_length`-year 
period, up to 2050.
The whole woodland grows to an age of `r rotn_length` years over this time.
From Figure 1, the carbon sequestration in woodland over this period is
`r round(S_c_25y, 2)` kg CO$_2$ / m$^2$, so a mean uptake rate of 
`r round(f_c_mean, 2)` kg CO$_2$ / m$^2$ / year (dashed line in Figure 2). 
The emissions from one cow is a constant `r e_cow` kg CO$_2$ / year.
So, to offset the `r rotn_length`-year emissions from one cow, you would need to plant 
`r round(required_affn, 0)` m$^2$ of new woodland (`r e_cow` kg CO₂ divided by 
`r round(f_c_mean, 2)` kg CO$_2$/m$^2$).

In some ways, the area the cattle occupy does not matter - you just need to 
plant `r round(required_affn, 0)` m$^2$ of new woodland per head of cattle. 
However, if you want to express the required afforestation as a fraction of the 
land area used for grazing, we need to specify the cattle stocking density.
A typical stocking density for conservation grazing is ~0.25 LU/ha = 1 cow per 4 
hectares = 1 cow per 40,000 m$^2$.
The required `r round(required_affn, 0)` m$^2$ of woodland would constitute 
`r round(required_affn, 0) / 40000 * 100` % of the grazed area.
A more typical stocking density for conventional grazing is ~1 LU/ha = 1 cow per 
10,000 m$^2$. The required `r round(required_affn, 0)` m$^2$ of woodland would constitute 
`r round(required_affn, 0) / 10000 * 100` % of the grazed area.

### Q1b. What area of land would need to be planted with trees *each year* to offset the methane emissions from a single cow over the next `r rotn_length` years (i.e. so as to meet the Net Zero 2050 target for that single cow)?
```{r, echo = FALSE, warning = FALSE, message = FALSE}
f_c_mean_w <- dt[time == rotn_length, f_c_mean_w]
required_affn <- e_cow / f_c_mean_w  # m2 of woodland needed to offset one cow's emissions per year
required_affn_1b <- required_affn
```
In this scenario, we consider the total area of woodland that would need to be planted
to offset the methane emissions from a single cow over the `r rotn_length`-year 
period 2025-2050, if the woodland planting was on-going with an equal area 
planted each year.
Only 1/`r rotn_length`th of the woodland grows to an age of `r rotn_length` 
years over this time, so the mean uptake rate is much lower - the dotted line in
Figure 2. The mean carbon sequestration across the mixed age woodland over this 
period is 
`r round(f_c_mean_w, 2)` kg CO$_2$ / m$^2$ / year. 
The emissions from one cow is still a constant `r e_cow` kg CO$_2$ / year.
So, to offset the `r rotn_length`-year emissions from one cow, you would need to plant 
`r round(required_affn, 0)` m$^2$ of new woodland (`r e_cow` kg CO₂ divided by 
`r round(f_c_mean_w, 2)` kg CO$_2$/m$^2$).

At a typical stocking density for conservation grazing is ~0.25 LU/ha = 
1 cow per 40,000 m$^2$, the required `r round(required_affn, 0)` m$^2$ of 
woodland would constitute 
`r round(required_affn, 0) / 40000 * 100` % of the grazed area.
At the more typical stocking density for conventional grazing of 1 LU/ha = 1 cow per 
10,000 m$^2$, the required `r round(required_affn, 0)` m$^2$ of woodland would 
constitute 
`r round(required_affn, 0) / 10000 * 100` % of the grazed area.

### Q2a. What area of land would need to be planted with trees *in 2025* to offset the methane emissions from a single cow, over the next 100 years?
Here, we repeat the same calculations but consider a longer time period where 
the woodland is assumed to grow to 100 years old. 
This makes a difference because tree growth is slow initially, 
faster from ~15-30 years, and then tails off to zero. The exact shape of the 
curve depends on the species.
```{r, echo = FALSE, warning = FALSE, message = FALSE}
rotn_length <- 100
# f_c_mean <- mean(dt$S_c[1:rotn_length]) # gives same answer
S_c_50y <- dt[time == rotn_length, S_c]
f_c_mean <- dt[time == rotn_length, f_c_mean]
f_c_mean <- dt[time == rotn_length, S_c / time]
f_c_mean <- set_units(f_c_mean, kg / m^2 / yr)
required_affn <- e_cow / f_c_mean  # m2 of woodland needed to offset one cow's emissions per year
required_affn_2a <- required_affn
```

In this scenario, we consider the area of woodland planted in 2025 needed to 
offset the methane emissions from a single cow over the following `r rotn_length`-year 
period, up to 2075.
The whole woodland grows to an age of `r rotn_length` years over this time.
From Figure 1, the carbon sequestration in woodland over this period is
`r round(S_c_50y, 2)` kg CO$_2$ / m$^2$, so a mean uptake rate of 
`r round(f_c_mean, 2)` kg CO$_2$ / m$^2$ / year (From Figure 2). 
The emissions from one cow is a constant `r e_cow` kg CO$_2$ / year.
So, to offset the `r rotn_length`-year emissions from one cow, you would need to plant 
`r round(required_affn, 0)` m$^2$ of new woodland (`r e_cow` kg CO₂ divided by 
`r round(f_c_mean, 2)` kg CO$_2$/m$^2$).

To express the required afforestation as a fraction of the 
land area used for grazing, we specify the same cattle stocking densities as before.
At a typical stocking density for conservation grazing of ~0.25 LU/ha 
= 1 cow per 40,000 m$^2$, the required `r round(required_affn, 0)` m$^2$ of 
woodland would constitute 
`r round(required_affn, 0) / 40000 * 100` % of the grazed area.
At the more typical stocking density for conventional grazing of ~1 LU/ha 
= 1 cow per 10,000 m$^2$, the required `r round(required_affn, 0)` m$^2$ of 
woodland would constitute 
`r round(required_affn, 0) / 10000 * 100` % of the grazed area.

### Q2b. What area of land would need to be planted with trees *each year* to offset the methane emissions from a single cow over the next `r rotn_length` years
```{r, echo = FALSE, warning = FALSE, message = FALSE}
f_c_mean_w <- dt[time == rotn_length, f_c_mean_w]
required_affn <- e_cow / f_c_mean_w  # m2 of woodland needed to offset one cow's emissions per year
required_affn_2b <- required_affn
```

In this scenario, we consider the total area of woodland that would need to be planted
to offset the methane emissions from a single cow over the `r rotn_length`-year 
period 2025-2125, if the woodland planting was on-going with an equal area 
planted each year.
Only 1/`r rotn_length`th of the woodland grows to an age of `r rotn_length` 
years over this time, and a bigger area is in the peak growth stage, so the mean 
uptake rate is higher - the dotted line in
Figure 2. The mean carbon sequestration across the mixed age woodland over this 
period is 
`r round(f_c_mean_w, 2)` kg CO$_2$ / m$^2$ / year. 
The emissions from one cow is still a constant `r e_cow` kg CO$_2$ / year.
So, to offset the `r rotn_length`-year emissions from one cow, you would need to plant 
`r round(required_affn, 0)` m$^2$ of new woodland (`r e_cow` kg CO₂ divided by 
`r round(f_c_mean_w, 2)` kg CO$_2$/m$^2$).

At a typical stocking density for conservation grazing is ~0.25 LU/ha = 
1 cow per 40,000 m$^2$, the required `r round(required_affn, 0)` m$^2$ of 
woodland would constitute 
`r round(required_affn, 0) / 40000 * 100` % of the grazed area.
At the more typical stocking density for conventional grazing of 1 LU/ha = 1 cow per 
10,000 m$^2$, the required `r round(required_affn, 0)` m$^2$ of woodland would 
constitute 
`r round(required_affn, 0) / 10000 * 100` % of the grazed area.

### National-scale context
To put these numbers in a national-scale context, we need to consider the UK
population of cattle and the land area available.
```{r, echo = FALSE, warning = FALSE, message = FALSE}
area_uk <- set_units(243000, km^2)
current_planting_rate <- set_units(20660, ha)
target_planting_rate <- set_units(30000, ha)
current_planting_rate <- set_units(current_planting_rate, km^2)
target_planting_rate <- set_units(target_planting_rate, km^2)
```

To offset the UK cattle herd of ~9.5 million [Defra 2024](https://www.gov.uk/government/statistics/livestock-populations-in-the-united-kingdom/livestock-populations-in-the-united-kingdom-at-1-december--2), 
half dairy and half beef, we can look at the national area of woodland planting 
required, given the four scenarios. For comparison, the current UK annual new 
planting rate is `r current_planting_rate` km$^2$/yr in 2023/4 [FR 2024](https://www.forestresearch.gov.uk/tools-and-resources/statistics/publications/forestry-statistics/forestry-statistics-2024/2024-1-woodland-area-and-planting/).
The target planting rate is `r target_planting_rate` km$^2$/yr.

1a. To offset the UK cattle emissions up to 2050 with up-front planting, we would need to plant
`r round(set_units(required_affn_1a * n_cows, km^2), 0)` km$^2$ in 2025/6.
This is `r round(required_affn_1a * n_cows / current_planting_rate, 0)` times 
the current planting rate, and equivalent to 
`r round(set_units(required_affn_1a * n_cows, km^2) / area_uk * 100, 0)` %
of the total UK land area.

1b. To offset the UK cattle emissions up to 2050 with on-going annual planting, we would need to plant
`r round(set_units(required_affn_1b * n_cows / 25, km^2), 0)` km$^2$ each year 
until 2050.
This is `r round(required_affn_1b * n_cows / 25 / current_planting_rate, 0)` times 
the current planting rate, and would give a total area equivalent to 
`r round(set_units(required_affn_1b * n_cows, km^2) / area_uk * 100, 0)` %
of the total UK land area.

2a. If we consider a `r rotn_length`-year period for woodland growth, we would need to plant
`r round(set_units(required_affn_2a * n_cows, km^2), 0)` km$^2$ up-front in 
2025/6. This is `r round(required_affn_2a * n_cows / current_planting_rate, 0)` times 
the current planting rate, and equivalent to 
`r round(set_units(required_affn_2a * n_cows, km^2) / area_uk * 100, 0)` %
of the total UK land area.

2b. With on-going annual planting over `r rotn_length` years, we would need to plant
`r round(set_units(required_affn_2b * n_cows / rotn_length, km^2), 0)` km$^2$ each year 
until 2075.
This is `r round(required_affn_2b * n_cows / rotn_length / current_planting_rate, 2)` times 
the current planting rate, and would give a total area equivalent to 
`r round(set_units(required_affn_2b * n_cows, km^2) / area_uk * 100, 0)` %
of the total UK land area.

### Q3. On a single farm with a fixed area, how long could grazing continue at a reasonable stocking density before the area of the land needed for woodland precludes grazing?
Less arbitrarily, we could assume that both the grazing and afforestation take 
place within the same fixed area. In this case, we start with 40000 m$^2$ of 
grassland per cow, and convert `r round(required_affn_2b, 0)` m$^2$ to woodland 
each year, assuming option 2b of an intended `r rotn_length`-year time horizon. In this scenario, 
grassland is gradually displaced by woodland over time, so the cattle stocking 
density increases until the whole area is covered by woodland after 
`r round(40000 / required_affn_2b, 0)` years.

### Conclusions
Comparing methane emissions from cows and CO$_2$ uptake by trees is not straightforward but a few points are clear.

- Carbon sequestration from planted woodland is transient (it occurs over a few decades and then declines to zero), whilst methane emissions per cow are constant.
- Because of this, in the long term we need to keep continually planting new woodland to offset constant emissions.
- To compare like-with-like, we need to specify a time period and a tree planting schedule (e.g. all-at-once or continuous).
- To offset emissions from the current UK cattle herd over the next 25 years, in time to meet the Net Zero 2050 target, we need to plant an additional 
`r round(set_units(required_affn_1b * n_cows, km^2) / area_uk * 100, 0)` % of 
the total UK land area, requiring 
`r round(required_affn_1b * n_cows / 25 / current_planting_rate, 0)`
times the current planting rate for 25 years (option 1b). Planting the required area up-front in 2025/26 (`r round(set_units(required_affn_1a * n_cows, km^2), 0)` km$^2$, 
option 1a) is not practically feasible. Considering on-going annual planting 
over 100 years makes this feasible within the current planting rate, but delays 
reaching a net zero balance until 2125.

In general, the uptake capacity of woodland planting on the scale feasible in 
the UK is small compared to the size of the emissions from the UK cattle population.
At the current planting rate, we are sequestering 
`r round(f_c_mean_w * current_planting_rate / (e_cow * n_cows) * 100, 3)` %
of the emissions from the UK cattle population
(`r round(set_units(f_c_mean_w * current_planting_rate, Tg/yr), 4)` Tg CO$_2$/yr, 
compared with emissions from the UK cattle population of
`r round(set_units(e_cow * n_cows, Tg/yr), 2)` Tg CO$_2$-e/yr).

<!-- As an independent check on these values, the UK GHG [inventory](https://www.gov.uk/government/statistics/final-uk-greenhouse-gas-emissions-statistics-1990-to-2023)
reports methane emissions from the UK cattle population as
```{r, eval = FALSE}
e_cow_inv <- set_units(21.2, Mt/yr)
e_cow_inv <- set_units(e_cow_inv, Tg/yr)
set_units(e_cow_inv / 1875, kg/yr)
set_units(e_cow_inv / n_cows, kg/yr)
round(set_units(e_cow * n_cows, Tg/yr), 2)
```
The UK GHG inventory does not report the uptake by woodland planted in a 
specific year, but the 10-year mean for uptake by all woodland planted in the 
previous 20 years is
```{r, eval = FALSE}
f_c <- set_units(0.2, Mt/yr)
f_c <- set_units(f_c, Tg/yr)
f_c / current_planting_rate
set_units(f_c / current_planting_rate, kg/m^2/yr)
round(set_units(f_c / current_planting_rate, kg/m^2/yr), 2)
``` -->

<!-- 1. Estimate the rotation length on which any harvesting might occur, and use the
mean carbon stock over that period. You can then calculate the area of woodland
that has to be planted and maintained each year on that basis to offset the 
annual methane emissions from one cow. This is an equilibrium approach, as it is
based on a long-term average carbon stock.

2. Consider a fixed time interval over which grazing is to take place, and 
equate with the woodland carbon sequestration which can place over this same 
interval. This is a transient approach, as it considers the the immediate 
carbon sequestration as it progresses, rather than assuming anything about a 
long-term equilibrium.

3. Consider a fixed area of land which is used for both grazing and woodland. 
Here the question becomes how long can grazing be sustained before the area is
completely converted to woodland (or some livestock stocking density threshold 
exceeded). This is also a transient approach, but makes no assumption about 
additional land being available elsewhere to be afforested. -->
